---
title: "Utility trips methodology"
date: "Last updated: `r Sys.Date()`"
author: "Joey Talbot and Robin Lovelace"
format: gfm
execute: 
  echo: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
```

```{r}
#| include: false
#| label: import-nts
u = "https://assets.publishing.service.gov.uk/media/64e8b433635870000d1dbfbc/nts0403.ods"
f = basename(u)
if (!file.exists(f)) download.file(u, f)
readODS::ods_sheets(f)
nts0403_trips_excl_sw = readODS::read_ods(f, sheet = "NTS0403b_trips_excl_sw", skip = 5)
names(nts0403_trips_excl_sw)
#  [1] "Year"                                     
#  [2] "Commuting"                                
#  [3] "Business"                                 
#  [4] "Education"                                
#  [5] "Escort education"                         
#  [6] "Shopping"                                 
#  [7] "Other escort"                             
#  [8] "Personal business"                        
#  [9] "Visiting friends at private home"         
# [10] "Visiting friends elsewhere"               
# [11] "Entertainment or public activity"         
# [12] "Sport: participate"                       
# [13] "Holiday: base"                            
# [14] "Day trip"                                 
# [15] "Other including just walk"                
# [16] "All purposes"                             
# [17] "Unweighted sample size: individuals"      
# [18] "Unweighted sample size: trips (thousands)"
nts0403_length_raw = readODS::read_ods(f, sheet = "NTS0403e_trip_length", skip = 5)
names(nts0403_length_raw)
#  [1] "Year"                                     
#  [2] "Commuting"                                
#  [3] "Business"                                 
#  [4] "Education"                                
#  [5] "Escort education"                         
#  [6] "Shopping"                                 
#  [7] "Other escort"                             
#  [8] "Personal business"                        
#  [9] "Visiting friends at private home"         
# [10] "Visiting friends elsewhere"               
# [11] "Entertainment or public activity"         
# [12] "Sport: participate"                       
# [13] "Holiday: base"                            
# [14] "Day trip"                                 
# [15] "Other including just walk"                
# [16] "All purposes"                             
# [17] "Unweighted sample size: individuals"      
# [18] "Unweighted sample size: trips (thousands)"
```

## Trip numbers and purposes

We estimated trips for three everyday purposes (in addition to commuting and travel to school):

- shopping
- social (visiting friends and family)
- leisure

These are presented in a single network layer under the joint category of 'Other Everyday' trips.

We used data from England's National Travel Survey 2019 to estimate the total number of trips undertaken per person per day, as AADT. 
This gives an average of 953 trips per person per year, which equates to 2.61 trips per person per day. 
A trip is defined as a one-way course of travel from one place to another (https://www.gov.uk/government/statistics/national-travel-survey-2022-technical-report/national-travel-survey-2022-technical-report-glossary#trip).

Data on the number of trips per person per year is not available for Scotland after 2012 as far as we can tell.
The Scottish Household Survey only records travel on the day before the survey.
The National Travel Survey records travel patterns over an entire week.
However, for the ten year period from 2002/03 to 2011/12, the National Travel Survey recorded a mean of 995 trips per person per year by Scottish residents. 
In England, during 2002 to 2012, there were 1023 trips per person per year (https://assets.publishing.service.gov.uk/media/64e8b00063587000141dbfa6/nts0303.ods)

We assigned daily trips to trip purposes using the trip purpose percentage breakdown from the Scottish Household Survey, Table TD3 of the Transport and Travel in Scotland 2019 travel diary tables (after adjusting these percentages to remove the 'Go Home' category).
This results in 25.1% of trips being assigned for shopping, 11.7% for social, and 6.3% for leisure, compared with 23.3% for commuting (https://www.transport.gov.scot/media/51346/transport-and-travel-in-scotland-2019-travel-diary-tables.xlsx).

The roughly equivalent purposes in the National Travel Survey for 2022 are as follows:


```{r}
npt_nts_purpose_categories = tibble::tribble(
  ~`NPT purpose`, ~`NTS purpose`,
  "Shopping", "Shopping",
  "Social", "Visiting friends at private home",
  "Social", "Visiting friends elsewhere",
  "Leisure", "Entertainment or public activity",
  "Leisure", "Sport: participate"
)
npt_proportions = tibble::tribble(
  ~`NPT purpose`, ~Proportion,
  "Shopping", 0.251,
  "Social", 0.117,
  "Leisure", 0.063
)
# Get subset of NTS trip data from 2011 onwards:
nts0403_trips_percent = nts0403_trips_excl_sw |>
  select(Year:`All purposes`) |>
  # Convert to proportions of All purposes:
  mutate(across(-Year, ~ . / `All purposes`)) |>
  filter(Year > "2011") |>
  pivot_longer(-Year, names_to = "NTS purpose", values_to = "Proportion") |>
   filter(str_detect(`NTS purpose`, "Shop|Visit|Entertain"))
trips_percent_2022 = nts0403_trips_percent |>
  # Convert to % with scales::percent_format():
  filter(Year == "2022") |>
  select(`NTS purpose`, Proportion) |>
  left_join(npt_nts_purpose_categories, by = "NTS purpose")
trips_percent_2022 |>
  mutate(Proportion = scales::percent(Proportion, accuracy = 0.1)) |>
  knitr::kable()
```

Aggregating the NTS purposes to match the NPT purposes, we find that the mode split for the three everyday purposes in 2022 is as follows:

```{r}
trips_percent_2022 |>
  group_by(`NPT purpose`) |>
  summarise(`Proportion (NTS)` = sum(as.numeric(Proportion))) |>
  left_join(npt_proportions, by = "NPT purpose") |>
  transmute(
    `NPT purpose`,
    `Proportion (NTS)` = scales::percent(`Proportion (NTS)`, accuracy = 0.1),
    `Proportion (NPT)` = scales::percent(Proportion, accuracy = 0.1)
  ) |>
  knitr::kable()
```

The evolution of overall mode split from 2012 to 2022 in England is shown below:

```{r}
#| label: nts-trips-mode-split-time
# Line graph of trip purposes:
nts0403_trips_percent |>
  ggplot() +
  geom_line(aes(x = Year, y = Proportion, color = `NTS purpose`, group = `NTS purpose`)) +
  # x label at 45 degrees:
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # integer x axis labels:
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  # % y axis labels:
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, NA)) 
```

## Trip distances by purpose

Data on the distribution of trip distances (what % of trips are 0-2, 2-5 km etc) is important for the uptake model.
We took data from the National Travel Survey, shown below, as the basis of average trip lengths for the three everyday purposes, as shown in the table below.

```{r, include=FALSE}
nts0403_length = pivot_longer(nts0403_length_raw, -1, values_to = "Length (miles)", names_to = "NTS purpose") |>
  # Remove the space and everything after it:
    mutate(Year = str_remove(Year, " .+")) 
nts0403_length_subset = nts0403_length |>
  filter(Year > "2011") |>
  filter(str_detect(`NTS purpose`, "Commut|Shop|Visit|Edu|Entertain")) 
```

```{r}
#| label: nts-trips-length-time
plot_NTS_categories = nts0403_length_subset |>
  ggplot() +
  geom_line(aes(x = Year, y = `Length (miles)`, color = `NTS purpose`, group = `NTS purpose`)) +
  # x label at 45 degrees:
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(c(0, NA))
lengths_npt = nts0403_length_subset |>
  filter(Year == "2022") |>
  left_join(npt_nts_purpose_categories, by = "NTS purpose") |>
  mutate(`NPT purpose` = case_when(
    is.na(`NPT purpose`) ~ `NTS purpose`,
    TRUE ~ `NPT purpose`
  ))
# Join with % trips:
lengths_npt = left_join(
  lengths_npt,
  trips_percent_2022 |> select(`NTS purpose`, Proportion),
  by = "NTS purpose"
) |>
  mutate(Proportion = case_when(
    is.na(Proportion) ~ 1,
    TRUE ~ `Length (miles)`
  )) |>
  group_by(`NPT purpose`) |>
  summarise(`Length (miles)` = weighted.mean(`Length (miles)`, Proportion)) 
lengths_npt |>
  mutate(`Length (miles)` = scales::number(`Length (miles)`, accuracy = 0.1)) |>
  knitr::kable()
```

## Mode share

The mode shares were estimated using the mean mode shares from the Scottish Household Survey travel diaries in Table TD2 of Transport and Travel in Scotland 2019
(https://www.transport.gov.scot/media/51346/transport-and-travel-in-scotland-2019-travel-diary-tables.xlsx).
This gives mode shares (for social and leisure trips) of 1.2% for cycling, 22.1% for walking, 65.2% for car, 9.3% for public transport, and 2.2% for taxi/other.

For shopping, we assumed that the cycle mode share was half of this, with the mode share for the other modes (car, public transport, walking and taxi/other) increased accordingly. 
This follows our detailed analysis of Scottish Household Survey travel diaries, which identified shopping as having a considerably lower cycle mode share than other trip purposes.
Shopping trips can be more difficult to cycle, since they require carrying larger volumes of luggage.

## Trip origins and destinations

Origins and destinations were assigned based on Intermediate Zones.
The trip origins were assumed to be people's homes.
The number of trips originating from each zone is therefore a function of the residential population of the zone, multiplied by the AADT for the given trip purpose.

For shopping and leisure trips, we used Ordnance Survey Points of Interest (POIs) to identify potential trip destinations. 
All of these POIs were assigned to the nearest point on a 500m grid, giving a density index for each grid point.
The grid was clipped to avoid any destination points falling offshore.

The POIs for shopping comprised any location classed as 'Retail.' 
For leisure, they comprised locations classed as 'Sport and Entertainment.' 
This includes gyms, swimming pools, sports centres, nightclubs, cinemas, theatres, casinos and similar venues.
We chose this category because it matches the Scottish Household Survey 'Sport/Entertainment' trip purpose.  

For social trips, the destinations were people's homes, so the number of trips arriving in each zone was a function of the zone's residential population. 

## Spatial interaction model

We used a spatial interaction model to pair trip origins and destinations and thus create desire lines.
These are assigned according to a distance decay equation, which has been derived from distances of travel to work.
This may introduce bias if in reality typical trip distances for these other everyday purposes are different from the trip distances for commuting.

See NTS0403e for mean trip lengths by purpose
- shopping
- visiting friends at private home
- entertainment or public activity

## Disaggregating and filtering desire lines

The resulting desire lines are put through a jittering process.
In cases where the trip origins or destinations were assigned based on Intermediate Zone populations, we disaggregate these desire lines and assign them to randomised points on the road network of the zone, rather than using a single point for each zone. 
The disaggregation threshold is 100.

We then filter the desire lines, excluding those with a Euclidean distance of <500m (below this distance people are likely to walk rather than cycle) or >5km (to reduce the computational time of the routing).

## Routing

Finally, the desire lines are routed on the road network, using the CycleStreets routing algorithms for fast and quiet routes.
Individual routes are then combined into a route network.

